Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-27 paper.tex
Grammar Mega-Pass

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On an FPG-lacking devices---Apple M1 Max and M3 ()---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPGPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained only by bandwidth and memory latency. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward progress guarantee}; without this guarantee, it risks indefinite starvation.

@@

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.

However, certain aspects fall outside the scope of this work. \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain speed-of-light performance. \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.

@@

\subsubsection{Inter-Workgroup Barriers}
Parallel programs in general require synchronization between multiple threads (those that don't are considered embarrassingly parallel). A standard technique for synchronization is \emph{barriers.} GPUs provide a limited palette of barrier primitives. At the coarsest granularity, \emph{pipeline barriers} fully synchronize both control and memory between multiple kernel launches. However, pipeline barriers are extremely expensive and efficient algorithms strictly minimize their use. At a much finer granularity, \emph{workgroup barriers} provide reasonably efficient synchronization between the threads in a workgroup. However, synchronization between workgroups is more limited. Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2. Because of its non-portability, it is also absent from the WebGPU specification. Even when available, the use of such memory barriers causes performance issues as they are usually implemented as a full cache flush.

@@

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{Horn, Hensley, Sengupta2006, Gress} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.

@@

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by storing only the work tile reduction in global memory during the first phase and performing a redundant scan in the third, it eliminates $n$ global data movement while preserving the three-phase structure:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

@@

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches and pipeline barriers as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barriers to implicit synchronization via access order. As a result, each element is read and written exactly once, achieving the theoretical minimum of $O(2n)$.

In any \emph{Chained-Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a \emph{forward progress guarantee} of the GPU's execution model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.

@@

\subsubsection{Chained-Scan, Decoupled-Lookback}
Introduced by Merrill and Garland, the \emph{Decoupled-Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

Although \emph{Decoupled-Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces negligible overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled-Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream-Scan}.

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design. By eliding the barrier-idiom, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.

@@

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically defined and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their findings demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior assumptions~\cite{sorensen2018}.

@@

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing shows that, at best, running \emph{Chained-Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled-Lookback (PLACEHOLDER DATA)}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-Then-Scan}, sacrificing performance for broader compatibility.

\subsection{Earlier Attempts at Portability}
Our first attempt \raph{we'll add the cite post-review} at implementing a \emph{Decoupled-Lookback} scan without relying on FPG was \emph{Scalar Fallback}. Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile. This eliminates the possibility of deadlock because the number of such operations is bouded; the wait will eventually terminate, possibly at the expense of doing redundant work.

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.

@@

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations. As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

@@

\subsection{Decoupled-Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback. Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\varepsilon$. State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions. To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models~\cite{Alglave0, MCMutants}. The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
\end{table}

@@

\subsubsection{Same-API Performance Versus Baseline Measures}
We next compare \emph{DF} against \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique. This evaluation is conducted using Dawn across the full range of tested devices.

>>>>

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPGPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward progress guarantee}; without this guarantee, it risks indefinite starvation.

@@

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.

However, certain aspects fall outside the scope of this work. \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain speed-of-light performance. \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained Scan}.

@@

\subsubsection{Inter-Workgroup Barriers}
Parallel programs in general require synchronization between multiple threads (those that don't are considered embarrassingly parallel). A standard technique for synchronization is \emph{barriers.} GPUs provide a limited palette of barrier primitives. At the coarsest granularity, \emph{pipeline barriers} fully synchronize both control and memory between multiple kernel launches. However, pipeline barriers are extremely expensive and efficient algorithms strictly minimize their use. At a much finer granularity, \emph{workgroup barriers} provide reasonably efficient synchronization between the threads in a workgroup. However, synchronization between workgroups is more limited. Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2. Because of its non-portability, it is also absent from the WebGPU specification. Even when available, the use of such memory barriers causes performance issues as they are usually implemented as a full cache flush.

@@

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{Horn, Hensley, Sengupta2006, Gress} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.

@@

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-then-Propagate} strategy called \emph{Reduce-then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by storing only the work tile reduction in global memory during the first phase and performing a redundant scan in the third, it eliminates $n$ global data movement while preserving the three-phase structure:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

@@

\subsubsection{Chained Scan, StreamScan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{StreamScan}, the key innovation of the \emph{Chained Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches and pipeline barriers as inter-workgroup synchronization points, \emph{Chained Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barriers to implicit synchronization via access order. As a result, each element is read and written exactly once, achieving the theoretical minimum of $O(2n)$.

In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a \emph{forward progress guarantee} of the GPU's execution model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.

@@

\subsubsection{Chained Scan, Decoupled Lookback}
Introduced by Merrill and Garland, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

Although \emph{Decoupled Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces negligible overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream Scan}.

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce-then-Scan} represent two contrasting paradigms in scan design. By eliding the barrier-idiom, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler. In contrast, \emph{Reduce-then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.

@@

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not. Without forward progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Why does \emph{Chained Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically defined and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their findings demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior assumptions~\cite{sorensen2018}.

@@

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing shows that, at best, running \emph{Chained Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled Lookback (PLACEHOLDER DATA)}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

\subsection{Earlier Attempts at Portability}
Our first attempt \raph{we'll add the cite post-review} at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}. Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile. This eliminates the possibility of deadlock because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce-then-Scan}.

@@

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations. As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

@@

\subsection{Decoupled-Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback. Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\varepsilon$. State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions. To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models~\cite{Alglave0, MCMutants}. The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
\end{table}

@@

\subsubsection{Same-API Performance Versus Baseline Measures}
We next compare \emph{DF} against \emph{Reduce-then-Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique. This evaluation is conducted using Dawn across the full range of tested devices.
